---
title: "Reproducible Research: Peer Assessment 1"
author: "Fabian Camilo Nieto"
date: "July 12, 2017"
output: 
  html_document:
    keep_md: true
---

This report is the submition for the course [Reproducible Research](https://www.coursera.org/learn/reproducible-research) project 1 week2.

## First Step Fork/Clone Repository
The [original repository](http://github.com/rdpeng/RepData_PeerAssessment1) was fork in my own repository: [ReproducibleResearchProjectWeek2](https://github.com/fabiancnieto/ReproducibleResearchProjectWeek2)

## Loading and preprocessing the data
The following R additional packages are used:
- dplyr
- lattice
```{r, echo=TRUE, include=FALSE}
library(dplyr)
library(lattice)
```
Set the working directory
```{r, echo=TRUE}
setwd("/home/fnieto/Documents/R-Programming-Training/datasciencecoursera/Reproducible_Research/Week2/Project")
```
Load the "activity.csv" file into the activityData R object
```{r, echo=TRUE}
activityData <-
    read.csv("activity.csv",
    stringsAsFactors = FALSE,
    na.strings = "NA")
    activityDf <- tbl_df(activityData)
    str(activityDf)
```
Notice that the "date" column has chr class, for fix that we should assign the proper class (type of vairable) for "date"
```{r, echo=TRUE}
activityDf <- activityDf %>% mutate(date = as.Date(date))
str(activityDf)
```
## What is mean total number of steps taken per day?
Group by the "date" variable the data frame, ignoring the NA values for the variable "steps"
```{r, echo=TRUE}
byDayActivity <-
    activityDf %>% filter(!is.na(steps)) %>% group_by(date)
stepsByDayActivity <-
    byDayActivity %>% summarize(stepsByDay = sum(steps), meanByDay = mean(steps))
```
Make a histogram of the total number of the steps take each day
```{r, echo=TRUE}
attach(stepsByDayActivity)
histogram(stepsByDay,
          main = "Total Steps per Day",
          xlab = "Total steps",
          ylab = "Day number")
```

**Mean** and **Median** of the total steps taken per day
```{r, echo=TRUE}
stepsByDayActivity %>% summarize(Mean = mean(stepsByDay), Median = median(stepsByDay))
```

## What is the average daily activity pattern?
Gruop by "interval"" variable and get the mean of the steps per "internal"
```{r, echo=TRUE}
by5MinInterval <-
    activityDf %>% filter(!is.na(steps)) %>% group_by(interval) %>% summarize(avgSteps = mean(steps))
by5MinInterval
```
Time series plot Average Dailt Activity Pattern
```{r, echo=TRUE}
attach(by5MinInterval)
xyplot(
    avgSteps ~ interval,
    main = "Average Daily Activity",
    xlab = "5 Min Interval",
    ylab = "Avg Steps",
    grid = TRUE,
    type = "l"
    )
```

Maximum number of the steps 5 minute interval 
```{r, echo=TRUE}
topInterval <-
    as.character(by5MinInterval %>% arrange(desc(avgSteps)) %>% filter(row_number() == 1) %>% select(interval))
```
Maximum number of the steps 5 minute interval is: `r topInterval` => (topInterval)

## Imputing missing values
```{r, echo=TRUE}
countNas <-
    as.character(activityDf %>% filter(is.na(steps)) %>% summarize(n()))
```
The total number of missing values are: `r countNas` => (countNas)

Calculate the mean of the steps
```{r, echo=TRUE}
meanStepsByDay <-
    activityDf %>% summarize(meanByDay = mean(steps, na.rm = TRUE))
```

Subset the original data frame with the observations with missing values for steps variable
```{r, echo=TRUE}
byDayNas <-
    activityDf %>% filter(is.na(steps)) %>% group_by(date) %>% mutate(steps = (as.numeric(meanStepsByDay)))
```

New data set with missing values filled in mergin the byDayActivity and byDayNas data frames
```{r, echo=TRUE}
byDayActivities <-
    rbind(byDayActivity, byDayNas)
dimOriginal <- dim(activityDf)
dimMergeDf <- dim(byDayActivities)
```
The dimension of the original data frame is: **`r dimOriginal`** and for the merge data frame is: **`r dimMergeDf`**

Calculate the total number of steps per day
```{r, echo=TRUE}
stepsByDayActivities <-
    byDayActivities %>% summarize(stepsByDay = sum(steps), meanByDay = mean(steps))
```
Make a histogram of the total number of the steps take each day
```{r, echo=TRUE}
attach(stepsByDayActivities)
histogram(stepsByDay,
          main = "Total Steps per Day",
          xlab = "Total steps",
          ylab = "Day number")
```

**Mean** and **Median** of the total steps taken per day
```{r, echo=TRUE}
stepsByDayActivities %>% summarize(Mean = mean(stepsByDay), Median = median(stepsByDay))
```
Because the number of the observations were increased when added the missing values the total of the steps increased but the mean and median decreased.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable indicating whether a given date is "weekday" or "weekend"
```{r, echo=TRUE}
byDayActivities <-
    byDayActivities %>% mutate(weekday = format(date, "%a")) %>%
    mutate(dayType = ifelse((weekday == "Sat" |
    weekday == "Sun"), "weekend", "weekday"))
```
Gruop by "interval"" variable and get the mean of the steps per "internal"
```{r, echo=TRUE}
by5MinInterval <-
    byDayActivities %>% group_by(dayType, interval) %>% summarize(avgSteps = mean(steps))
by5MinInterval
```
Time series plot Average Daily Activity Pattern
```{r, echo=TRUE}
attach(by5MinInterval)
xyplot(
    avgSteps ~ interval|dayType,
    main = "Average Daily Activity",
    xlab = "5 Min Interval",
    ylab = "Avg number of Steps",
    grid = TRUE,
    type = "l"
    )
```